Report on the Third Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE3)

### 2015-11-07 WSSSPE3/report/wg_main_funding_spec_expert.tex
Americanizing (and more fixing) spelling, edits, notes in RSE main section

\subsubsection{Why it is important}

Research Software Engineers -- those who contribute to science and scholarship
through computer programming -- are an important part of the team
needed to deliver 21st century research. However, existing academic structures
and systems of funding do not effectively fund and sustain these skills.
The resulting high levels of turnover, and inappropriate incentivisation,
are a significant contributing factor to low levels of reliability and
readability observed in scientific software.

@@

A contributing factor to the problem is the devastating effect on an academic
publication record resulting from time in industry; such postings often
develop exactly the skills that research programmers need, yet returns to
university positions following an industry role are penalised by the current
structures. Retention of senior programmers is hard, because these people
are highly in demand by the economy. However,
people who have a PhD in science and enter industry,
may desire to return for diverse reasons, and should be welcomed back.

@@

It is hard for an individual PI in a University or College
to support dedicated research programming
resources, as the need for, and funding for, these activities is intermittent
within the research cycle. To sustain this capacity, therefore, it is
necessary to aggregate this work across multiple research groups.

One solution is to fund dedicated programming roles for major research software
projects at national laboratories or other non-educational institutions. This
solution is working well for many well-used scientific codebases.
However, much of the body of software for which this expertise is needed sits
within the longer tail of research in universities. Therefore, we argue that
institutions need to develop hybrid academic-technical roles for this capacity
who work with more than one PI, rather than the traditional RA role within a
single group. This could be centrally, perhaps within research computing
organisations which have traditionally supported university cyberinfrastructure,
library organisations, or research offices. Alternatively, these groups could
be organisationally closer to research groups, sitting within academic
departments. The most
effective model will vary from institution to institution,
but the mandate and ways of working should be similar.

@@

This might come from unversities themselves (this was the model that led to the
creation of the group in University College London), but more likely, seed
funding needs to come from research councils (as with the Research Software
Engineering Fellowship provided by the UK Engineering and Physical Sciences
Research Council). We therefore recommend that funding organisations consider
how they might provide such seed funding.

@@

We finally noted that the institutional case for such groups is made easier
by having successful examples to point to. In the UK, a collective effort to
identify this community, with individuals clearly stating ``I'm a research
software engineer'', has been important to the campaign. It will be useful to
the global effort to similarly identify emerging research software organisations,
and also, importantly, to identify longer-running research software groups,
which have in some cases had a long running \emph{sui-generis} existence, but which
now can be identified as part of a wider solution.

@@

We will therefore seek to gather a list of groups which are prepared to
self-identify as research software engineering groups. We will collect information
as to the organisational models under which these groups function, and how they
are funded.

@@

http://www.rse.ac.uk/groups

>>>>

\katznote{I don't think this section title is exactly right - are there some other options?  Also, the title of Appendix~\ref{sec:appendix_funding_spec_expert} should match, whatever is chosen.  Maybe `Funding Research Programmer Expertise'?}

%\subsubsection{Why it is important}

Research Software Engineers -- those who contribute to science and scholarship
through computer programming -- are an important part of the team
needed to deliver 21st century research. However, existing academic structures
and systems of funding do not effectively fund and sustain these skills.
The resulting high levels of turnover, and inappropriate incentives,
are a significant contributing factor to low levels of reliability and
readability observed in scientific software.

@@

A contributing factor to the problem is the devastating effect on an academic
publication record resulting from time in industry; such postings often
develop exactly the skills that research programmers need, yet returns to
university positions following an industry role are penalized by the current
structures. Retention of senior programmers is hard, because these people
are highly in demand by the economy. However,
people who have a PhD in science and enter industry,
may desire to return for diverse reasons, and should be welcomed back.

@@

It is hard for an individual PI in a university or college
to support dedicated research programming
resources, as the need for, and funding for, these activities is intermittent
within the research cycle. To sustain this capacity, therefore, it is
necessary to aggregate this work across multiple research groups.

One solution is to fund dedicated programming roles for major research software
projects at national laboratories or other non-educational institutions. This
solution is working well for many well-used scientific codebases.
However, much of the body of software for which this expertise is needed sits
within the longer tail of research in universities. Therefore, we argue that
institutions need to develop hybrid academic-technical roles for this capacity
who work with more than one PI, rather than the traditional RA role within a
single group. This could be centrally, perhaps within research computing
organizations which have traditionally supported university cyberinfrastructure,
library organizations, or research offices. Alternatively, these groups could
be organizationally closer to research groups, sitting within academic
departments. The most
effective model will vary from institution to institution,
but the mandate and ways of working should be similar.

@@

This might come from universities themselves (this was the model that led to the
creation of the group in University College London), but more likely, seed
funding needs to come from research councils (as with the Research Software
Engineering Fellowship provided by the UK Engineering and Physical Sciences
Research Council). We therefore recommend that funding organizations consider
how they might provide such seed funding.

@@

We finally noted that the institutional case for such groups is made easier
by having successful examples to point to. In the UK, a collective effort to
identify this community, with individuals clearly stating ``I'm a research
software engineer'', has been important to the campaign. It will be useful to
the global effort to similarly identify emerging research software organizations,
and also, importantly, to identify longer-running research software groups,
which have in some cases had a long running \emph{sui-generis} existence, but which
now can be identified as part of a wider solution.

For more detail on the discussion, see Appendix~\ref{sec:appendix_funding_spec_expert}.

@@

We will therefore seek to gather a list of groups which are prepared to
self-identify as research software engineering groups. We will collect information
as to the organizational models under which these groups function, and how they
are funded.

@@

To find more information about this group, or join it, see \url{http://www.rse.ac.uk/groups}.
